---
title: "Using the TimeSeriesAnalysis Package"
author: "Anthony G. Sena"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    toc: yes     
  html_document:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Using the TimeSeriesAnalysis Package}
  %\VignetteEncoding{UTF-8}    
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(TimeSeriesAnalysis)
library(dplyr)
```

# Overview

This document aims to showcase how to use the TimeSeriesAnalysis package for the sample data sets included in the package.

## Data set 1: Changes in use of HCQ in RA patients over time

Background: RA guidelines recommend using Methotrexate as first line therapy ASAP from 2008,
possibly ‘displacing’ hydroxychloroquine (HCQ).

- Period: 2000 – 2018

```{r}
outputFolder <- "E:/Timeseries/HCQ_In_RA_Example"
if (dir.exists(outputFolder)) {
  unlink(outputFolder, recursive = TRUE)
}
data(drugData)

# HCQ Use in RA Patients is cohortDefinitionId == 3
tsData <- drugData %>%
  filter(cohortDefinitionId == 3) %>%
  arrange(cohortStartDate) %>%
  select(cohortStartDate, subjectCount) %>%
  rename(eventDate = cohortStartDate,
         eventCount = subjectCount)

tsData
```

## Creating the model arguments

The following sections will describe how to define the various model fitting arguments for several use cases.

### Single Pre-specified change point – fixed (i.e. no estimated changepoint produced by model)

In this example, we'll create the arguments used to fit a model that has no estimated change points and instead provides a single pre-specified change point for evaluation.

```{r}
segArgs1 <- createSegmentedArgs(modelType = "linear",
                                psi = lubridate::as_date("2001-01-01"),
                                control = segmented::seg.control(it.max = 0,
                                                                 K = 1))

```

### Single prespecified changepoint – not fixed (i.e. an estimated changepoint produced by model which may or may no coincide with pre-specified change point)

Now we'll create the arguments used to fit a model that has 1 estimated change points and a single pre-specified change point for evaluation.

```{r}
segArgs2 <- createSegmentedArgs(modelType = "linear",
                                npsi = 1,
                                fixed.psi = list(eventDate = lubridate::as_date("2001-01-01")))
```


### No prespecified changepoint – an estimated changepoint produced by model

Now we'll create the arguments used to fit a model with a single estimated change point

```{r}
segArgs3 <- createSegmentedArgs(modelType = "linear")
```

### Single prespecified changepoint – and multiple estimated changepoints produced by model

Now we'll create the arguments used to fit a model with a single pre-specified changepoint and 2 estimated changepoints.

```{r}
segArgs4 <- createSegmentedArgs(modelType = "linear",
                                npsi = 2,
                                fixed.psi = list(eventDate = lubridate::as_date("2001-01-01")))
```

### Bayesian Online Change point detection

```{r}
ocpArgs <- TimeSeriesAnalysis::createOcpArgs()
```

## Fitting the models

Next we'll provide some code for using the arguments and data above to fit the models.

```{r}
# Create the full set of analyses
tsAnalysis1 <- createTsAnalysis(analysisId = 1,
                                description = "Single fixed, pre-specified change point",
                                tsArgs = segArgs1)
tsAnalysis2 <- createTsAnalysis(analysisId = 2,
                               description = "1 pre-specified, 1 estimated changepoint",
                               tsArgs = segArgs2)
tsAnalysis3 <- createTsAnalysis(analysisId = 3,
                                description = "Single estimated changepoint",
                                tsArgs = segArgs3)
tsAnalysis4 <- createTsAnalysis(analysisId = 4,
                                description = "1 pre-specified, 2 estimated changepoints",
                                tsArgs = segArgs4)
tsAnalysis5 <- createTsAnalysis(analysisId = 5,
                                description = "Bayesian Online Change point detection",
                                tsArgs = ocpArgs)

tsAnalysisList <- list(tsAnalysis1, tsAnalysis2, tsAnalysis3, tsAnalysis4, tsAnalysis5)

# Run the analysis
runTsAnalyses(tsData = tsData,
              tsDataId = 1, # A unique identifier for the data set
              outputFolder = outputFolder,
              tsAnalysisList = tsAnalysisList)
```

## Inspecting the results

Next we can use the package to plot the original time series and the model estimated change points.

### Single Pre-specified change point – fixed (i.e. no estimated changepoint produced by model)

```{r}
m1 <- readRDS(file = file.path(outputFolder, "Analysis1/ts_d1.rds"))
TimeSeriesAnalysis::plotSegmented(m1$tsData, m1$model, plotSubtitle = tsAnalysisList[[1]]$description)
```

### Single prespecified changepoint – not fixed (i.e. an estimated changepoint produced by model which may or may no coincide with pre-specified change point)

```{r}
m2 <- readRDS(file = file.path(outputFolder, "Analysis2/ts_d1.rds"))
TimeSeriesAnalysis::plotSegmented(m2$tsData, m2$model, plotSubtitle = tsAnalysisList[[2]]$description)
```

### No prespecified changepoint – an estimated changepoint produced by model

```{r}
m3 <- readRDS(file = file.path(outputFolder, "Analysis3/ts_d1.rds"))
TimeSeriesAnalysis::plotSegmented(m3$tsData, m3$model, plotSubtitle = tsAnalysisList[[3]]$description)
```

### Single prespecified changepoint – and multiple estimated changepoints produced by model

```{r}
m4 <- readRDS(file = file.path(outputFolder, "Analysis4/ts_d1.rds"))
TimeSeriesAnalysis::plotSegmented(m4$tsData, m4$model, plotSubtitle = tsAnalysisList[[4]]$description)
```

### Bayesian Online Change point detection
```{r}
m5 <- readRDS(file = file.path(outputFolder, "Analysis5/ts_d1.rds"))
TimeSeriesAnalysis::plotOcp(m5$tsData, m5$model, plotSubtitle = tsAnalysisList[[5]]$description)
```

## Data Set 2: HCQ for Covid-19 in 2020

Background:
- HCQ received emergency approval by FDA in March
- Regulatory action followed in late April due to CV safety issues
- Data: HCQ for Covid-19 treatment
- Period: Jan – Oct 2020


```{r}
outputFolder <- "E:/Timeseries/HCQ_In_COVID"
if (dir.exists(outputFolder)) {
  unlink(outputFolder, recursive = TRUE)
}

# HCQ for Covid-19 in 2020 is cohortDefinitionId == 1
tsData <- drugData %>%
  filter(cohortDefinitionId == 1) %>%
  arrange(cohortStartDate) %>%
  select(cohortStartDate, subjectCount) %>%
  rename(eventDate = cohortStartDate,
         eventCount = subjectCount)

tsData
```

### Single Pre-specified change point – fixed (i.e. no estimated changepoint produced by model)


```{r}
segArgs1 <- createSegmentedArgs(modelType = "linear",
                                psi = lubridate::as_date("2020-03-01"),
                                control = segmented::seg.control(it.max = 0,
                                                                 K = 1))

```

### Single prespecified changepoint – not fixed (i.e. an estimated changepoint produced by model which may or may no coincide with pre-specified change point)

```{r}
segArgs2 <- createSegmentedArgs(modelType = "linear",
                                npsi = 1,
                                fixed.psi = list(eventDate = lubridate::as_date("2020-03-01")))
```


### No prespecified changepoint – an estimated changepoint produced by model

Now we'll create the arguments used to fit a model with a single estimated change point

```{r}
segArgs3 <- createSegmentedArgs(modelType = "linear")
```

### Single prespecified changepoint – and multiple estimated changepoints produced by model

Now we'll create the arguments used to fit a model with a single pre-specified changepoint and 2 estimated changepoints.

```{r}
segArgs4 <- createSegmentedArgs(modelType = "linear",
                                npsi = 2,
                                fixed.psi = list(eventDate = lubridate::as_date("2020-03-01")))
```


## Fitting the models

Next we'll provide some code for using the arguments and data above to fit the models.

```{r}
# Create the full set of analyses
tsAnalysis1 <- createTsAnalysis(analysisId = 1,
                                description = "Single fixed, pre-specified change point",
                                tsArgs = segArgs1)
tsAnalysis2 <- createTsAnalysis(analysisId = 2,
                               description = "1 pre-specified, 1 estimated changepoint",
                               tsArgs = segArgs2)
tsAnalysis3 <- createTsAnalysis(analysisId = 3,
                                description = "Single estimated changepoint",
                                tsArgs = segArgs3)
tsAnalysis4 <- createTsAnalysis(analysisId = 4,
                                description = "1 pre-specified, 2 estimated changepoints",
                                tsArgs = segArgs4)
tsAnalysis5 <- createTsAnalysis(analysisId = 5,
                                description = "Bayesian Online Change point detection",
                                tsArgs = ocpArgs)

tsAnalysisList <- list(tsAnalysis1, tsAnalysis2, tsAnalysis3, tsAnalysis4, tsAnalysis5)

# Run the analysis
runTsAnalyses(tsData = tsData,
              tsDataId = 2, # A unique identifier for the data set
              outputFolder = outputFolder,
              tsAnalysisList = tsAnalysisList)
```


### Inspect the results

### Single Pre-specified change point – fixed (i.e. no estimated changepoint produced by model)

```{r}
m1 <- readRDS(file = file.path(outputFolder, "Analysis1/ts_d2.rds"))
TimeSeriesAnalysis::plotSegmented(m1$tsData, m1$model, plotSubtitle = tsAnalysisList[[1]]$description)
```

### Single prespecified changepoint – not fixed (i.e. an estimated changepoint produced by model which may or may no coincide with pre-specified change point)

```{r}
m2 <- readRDS(file = file.path(outputFolder, "Analysis2/ts_d2.rds"))
TimeSeriesAnalysis::plotSegmented(m2$tsData, m2$model, plotSubtitle = tsAnalysisList[[2]]$description)
```

### No prespecified changepoint – an estimated changepoint produced by model

```{r}
m3 <- readRDS(file = file.path(outputFolder, "Analysis3/ts_d2.rds"))
TimeSeriesAnalysis::plotSegmented(m3$tsData, m3$model, plotSubtitle = tsAnalysisList[[3]]$description)
```

### Single prespecified changepoint – and multiple estimated changepoints produced by model

```{r}
m4 <- readRDS(file = file.path(outputFolder, "Analysis4/ts_d2.rds"))
TimeSeriesAnalysis::plotSegmented(m4$tsData, m4$model, plotSubtitle = tsAnalysisList[[4]]$description)
```

### Bayesian Online Change point detection
```{r}
m5 <- readRDS(file = file.path(outputFolder, "Analysis5/ts_d2.rds"))
TimeSeriesAnalysis::plotOcp(m5$tsData, m5$model, plotSubtitle = tsAnalysisList[[5]]$description)
```


## Data Set 3: Dexamethasone for Covid-19 in 2020

Background:
- RECOVERY RCT report 30% reduction in mortality in late June 2020 with dexamethasone (DXM)
- Period: Jan – Oct 2020


```{r}
outputFolder <- "E:/Timeseries/DEX_In_COVID"
if (dir.exists(outputFolder)) {
  unlink(outputFolder, recursive = TRUE)
}

# DEX for Covid-19 in 2020 is cohortDefinitionId == 2
tsData <- drugData %>%
  filter(cohortDefinitionId == 2) %>%
  arrange(cohortStartDate) %>%
  select(cohortStartDate, subjectCount) %>%
  rename(eventDate = cohortStartDate,
         eventCount = subjectCount)

tsData
```

### Single Pre-specified change point – fixed (i.e. no estimated changepoint produced by model)


```{r}
segArgs1 <- createSegmentedArgs(modelType = "linear",
                                psi = lubridate::as_date("2020-06-01"),
                                control = segmented::seg.control(it.max = 0,
                                                                 K = 1))

```

### Single prespecified changepoint – not fixed (i.e. an estimated changepoint produced by model which may or may no coincide with pre-specified change point)

```{r}
segArgs2 <- createSegmentedArgs(modelType = "linear",
                                npsi = 1,
                                fixed.psi = list(eventDate = lubridate::as_date("2020-06-01")))
```


### No prespecified changepoint – an estimated changepoint produced by model

Now we'll create the arguments used to fit a model with a single estimated change point

```{r}
segArgs3 <- createSegmentedArgs(modelType = "linear")
```

### Single prespecified changepoint – and multiple estimated changepoints produced by model

Now we'll create the arguments used to fit a model with a single pre-specified changepoint and 2 estimated changepoints.

```{r}
segArgs4 <- createSegmentedArgs(modelType = "linear",
                                npsi = 2,
                                fixed.psi = list(eventDate = lubridate::as_date("2020-06-01")))
```


## Fitting the models

Next we'll provide some code for using the arguments and data above to fit the models.

```{r}
# Create the full set of analyses
tsAnalysis1 <- createTsAnalysis(analysisId = 1,
                                description = "Single fixed, pre-specified change point",
                                tsArgs = segArgs1)
tsAnalysis2 <- createTsAnalysis(analysisId = 2,
                               description = "1 pre-specified, 1 estimated changepoint",
                               tsArgs = segArgs2)
tsAnalysis3 <- createTsAnalysis(analysisId = 3,
                                description = "Single estimated changepoint",
                                tsArgs = segArgs3)
tsAnalysis4 <- createTsAnalysis(analysisId = 4,
                                description = "1 pre-specified, 2 estimated changepoints",
                                tsArgs = segArgs4)
tsAnalysis5 <- createTsAnalysis(analysisId = 5,
                                description = "Bayesian Online Change point detection",
                                tsArgs = ocpArgs)

tsAnalysisList <- list(tsAnalysis1, tsAnalysis2, tsAnalysis3, tsAnalysis4, tsAnalysis5)

# Run the analysis
runTsAnalyses(tsData = tsData,
              tsDataId = 3, # A unique identifier for the data set
              outputFolder = outputFolder,
              tsAnalysisList = tsAnalysisList)
```


### Inspect the results

### Single Pre-specified change point – fixed (i.e. no estimated changepoint produced by model)

```{r}
m1 <- readRDS(file = file.path(outputFolder, "Analysis1/ts_d3.rds"))
TimeSeriesAnalysis::plotSegmented(m1$tsData, m1$model, plotSubtitle = tsAnalysisList[[1]]$description)
```

### Single prespecified changepoint – not fixed (i.e. an estimated changepoint produced by model which may or may no coincide with pre-specified change point)

```{r}
m2 <- readRDS(file = file.path(outputFolder, "Analysis2/ts_d3.rds"))
TimeSeriesAnalysis::plotSegmented(m2$tsData, m2$model, plotSubtitle = tsAnalysisList[[2]]$description)
```

### No prespecified changepoint – an estimated changepoint produced by model

```{r}
m3 <- readRDS(file = file.path(outputFolder, "Analysis3/ts_d3.rds"))
TimeSeriesAnalysis::plotSegmented(m3$tsData, m3$model, plotSubtitle = tsAnalysisList[[3]]$description)
```

### Single prespecified changepoint – and multiple estimated changepoints produced by model

```{r}
m4 <- readRDS(file = file.path(outputFolder, "Analysis4/ts_d3.rds"))
TimeSeriesAnalysis::plotSegmented(m4$tsData, m4$model, plotSubtitle = tsAnalysisList[[4]]$description)
```

### Bayesian Online Change point detection
```{r}
m5 <- readRDS(file = file.path(outputFolder, "Analysis5/ts_d3.rds"))
TimeSeriesAnalysis::plotOcp(m5$tsData, m5$model, plotSubtitle = tsAnalysisList[[5]]$description)
```

